Cheminformatics: The Computer Science of Chemical Discovery

### 2011-11-12 paper.tex
Updated ontoogy section

While cheminformatics has a broad applicability, ranging from
agrochemical research and drug discovery to the design of novel
materials, we have chosen to use drug discovery as the context of this
article due to its central relevance for human wellbeing. Given that
the safety of drugs on the market is a key concern both for the
general audience as well as researchers in pharmaceutical companies,
the remainder of this article is structured around the concept of
``risk minimization'' in drug discovery -- the question of how to
minimize the chances of a small molecule failing, due to poor
physical, chemical or biological properties, in the various research
and development stages as a drug candidate, and how to successfully
employ cheminformatics methods in these efforts. As an example, a
molecule must be soluble and show a certain degree of bioavailability
to be considered as drug candidate. In cases where these properties
are poor, a cheminformatics approach can suggest replacement of
certain functional groups, that maintain potency but improve the
solubility and bioavailability.  Table~\ref{table:properties} (in the
Appendix) summarizes a number of properties a drug candidate must
satisfy to be considered therapeutically useful (such as being
appropriately soluble, exhibiting minimal toxicity and being selective
for the intended target) and sketches the role of cheminformatics at
each stage of drug development.
%
\subsection{Databases \& Ontologies}
\label{sec:databases}
%
Most cheminformatics applications rely on large databases of chemical structures,
their properties, and relationships to, \textit{e.g.}, biological targets.
Organizing and maintaining such databases, as well as searching and clustering similar structures together,
are essential  to enable many scientific applications. Yet, each one of
those areas poses its unique computer science challenges.

@@

%Fingerprints also facilitate the rapid calculation of quantitative
%structure-based similarity measures. An example measure of similarity is the
%\emph{Tanimoto coefficient}, calculated as the ratio $T(a,b) = \frac{c}{(a+b-c)}$,
%where $c$ is the count of bits on in the same position in both the two
%fingerprints, $a$ is the count of bits on in the first structure, and $b$ is the
%count of bits on in the second structure. Historically, this coefficient originated from the comparison of ecosystems
%around lakes, where a measure was needed to describe how similar those ecosystems are with respect to the animals,
%plants etc. they contain. However, it can be applied to every other system as well where features can be described in a
%binary manner. The Tanimoto coefficient varies in the
%range 0.0 -- 1.0, with a score of 1.0 indicating that the two structures are either very
%similar or even identical (i.e. their fingerprints are the same).
%
Structure-based similarity is however of limited use in grouping of
chemicals based on non-structural features, such as shared bioactivity
profiles. For this purpose, a large-scale and flexible classification
structure is needed for chemicals and their linked biological
information. One such answer to this need is provided in the form of
\emph{ontologies}.  An ontology is a formal specification of entities
and their relationships in a domain of interest. It is structured
around an underlying logical representation such as Description
Logics~\cite{baaderdl2007}. The logic-based representation allows
automated reasoning to derive new knowledge (inferences) from the
knowledge encoded, making the representation both compact and
powerful.  Ontologies are already in widespread use across the
biomedical domain, and in the cheminformatics field ontologies
relating to the chemical structures themselves and the protein targets
the molecules are active against are probably most common. A
similarity measure based on the information encoded in ontologies is
termed \emph{semantic similarity}, and such measures have already been
applied to enhance chemical classification in several problem areas
\cite{couto2010}. Many open questions still remain about the best way
to combine the information encoded in chemical structures with
chemical ontologies \cite{hastingsowled2010}.

>>>>

While cheminformatics has a broad applicability, ranging from
agrochemical research and drug discovery to the design of novel
materials, we have chosen to use drug discovery as the context of this
article due to its central relevance for human wellbeing. Given that
the safety of drugs on the market is a key concern both for the
general audience as well as researchers in pharmaceutical companies,
the remainder of this article is structured around the concept of
``risk minimization'' in drug discovery -- the question of how to
minimize the chances of a small molecule failing, due to poor
physical, chemical or biological properties, in the various research
and development stages as a drug candidate, and how to successfully
employ cheminformatics methods in these efforts. As an example, a
molecule must be soluble and show a certain degree of bioavailability
to be considered as drug candidate. In cases where these properties
are poor, a cheminformatics approach can suggest replacement of
certain functional groups, that maintain potency but improve the
solubility and bioavailability.  Table~\ref{table:properties} (in the
Appendix) summarizes a number of properties a drug candidate must
satisfy to be considered therapeutically useful (such as being
appropriately soluble, exhibiting minimal toxicity and being selective
for the intended target) and sketches the role of cheminformatics at
each stage of drug development.
%
\subsection{Representing and Searching Structures}
\label{sec:databases}
%
Most cheminformatics applications rely on large databases of chemical structures,
their properties, and relationships to, \textit{e.g.}, biological targets.
Organizing and maintaining such databases, as well as searching and clustering similar structures together,
are essential  to enable many scientific applications. Yet, each one of
those areas poses its unique computer science challenges.

@@

\subsection{Linking Molecules to Other Data}
\label{sec:prof-ident}

The preceding section focuses on the choice of representation of
chemical structures. With the profusion of databases that contain
details of small molecules, an emerging problem is the integration of
such databases. Such integration forms the basis of ``systems'' level
investigations of biological systems. A current challege in
cheminformatics is how to identify ``equivalent'' objects across data
sources. At one level, equivalent compounds can be identified via
canonical structure representations (InChI). But there are other
equivalences beyond structure. For example, compounds may bind to the
same or similar targets; they may be have similar biological function
(say, anti-neoplastic) and so on. To assess such higher level
characteristics, ontologies and semantic technologies have been
employed, to first, support annotations of chemical systems and
second, enable semantic analyses of chemical systems in conjunction
with other biological systems (such as pathways). A number of
challenges remain in this area - the use of ontologies allows one to
build a semntic network of concepts that links molecules to other
molecules as well as other objects such as targets, genes and
pathways. How does one evaluate similarities in such a heterogenous
network \cite{couto2010}? How does one integrate chemical structure
information with ontologies \cite{hastingsowled2010}? Integrating
small molecule data with other biological data types will be key to
acheiving a systems level understanding of small
molecules \cite{Oprea:2007fk}.

% Fingerprints also facilitate the rapid calculation of quantitative
% structure-based similarity measures. An example measure of
% similarity is the \emph{Tanimoto coefficient}, calculated as the
% ratio $T(a,b) = \frac{c}{(a+b-c)}$, where $c$ is the count of bits
% on in the same position in both the two fingerprints, $a$ is the
% count of bits on in the first structure, and $b$ is the count of
% bits on in the second structure. Historically, this coefficient
% originated from the comparison of ecosystems around lakes, where a
% measure was needed to describe how similar those ecosystems are with
% respect to the animals, plants etc. they contain. However, it can be
% applied to every other system as well where features can be
% described in a binary manner. The Tanimoto coefficient varies in the
% range 0.0 -- 1.0, with a score of 1.0 indicating that the two
% structures are either very similar or even identical (i.e. their
% fingerprints are the same).
%

% Structure-based similarity is however of limited use in grouping of
% chemicals based on non-structural features, such as shared bioactivity
% profiles. For this purpose, a large-scale and flexible classification
% structure is needed for chemicals and their linked biological
% information. One such answer to this need is provided in the form of
% \emph{ontologies}.  An ontology is a formal specification of entities
% and their relationships in a domain of interest. It is structured
% around an underlying logical representation such as Description
% Logics~\cite{baaderdl2007}. The logic-based representation allows
% automated reasoning to derive new knowledge (inferences) from the
% knowledge encoded, making the representation both compact and
% powerful.  Ontologies are already in widespread use across the
% biomedical domain, and in the cheminformatics field ontologies
% relating to the chemical structures themselves and the protein targets
% the molecules are active against are probably most common. A
% similarity measure based on the information encoded in ontologies is
% termed \emph{semantic similarity}, and such measures have already been
% applied to enhance chemical classification in several problem areas
% \cite{couto2010}. Many open questions still remain about the best way
% to combine the information encoded in chemical structures with
% chemical ontologies \cite{hastingsowled2010}.
